The function-sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-25 crelle-dictionary.tex
Fixed a few typos.

  Now let $(\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$
  and let $\rho : \W(G, \bg) \to \EEx$ be the character determined by $(\gqcs{L},\phi)$.
  Composing with the splitting $\Weil{} \to \W(G,\bg)$ yields an $\ell$-adic character
  $\rho_g : \Weil{} \to \EEx$, which is the same as the Trace of Frobenius defined in Section~\ref{ssec:Frob}, for every $\Fq$ rational point $g$ on $G$:
  $
  \rho_g(\Frob{}) =  \trFrob{\qcs{L}}(g).
  $
%
  On the other hand, we have already seen that $\trFrob{\qcs{L}} : G(\Fq) \to \EEx$
  is a group homomorphism. Since $G$ is an algebraic group over $\Fq$, it is a
  variety over $\Fq$ and thus $G(\Fq)$ is finite.
  Therefore $\trFrob{\qcs{L}}(g) = \rho_g(\Frob{})$ is a root of unity
  for every $g\in G(\Fq)$.  Since $\Weil{}$ is generated by
  $\Frob{}$ and $\rho_g : \Weil{} \to \EEx$ is
  a character, it follows that the image of $\rho_g$ is a finite group.
  Thus, $\rho_g$ extends to an $\ell$-adic character of $\Gal(\bFq/\Fq)$,
  which we will also denote $\rho_g$.
%
  We may now lift the $\ell$-adic character $\rho_g : \Gal(\bFq/\Fq) \to \EEx$
  to an $\ell$-adic character $\pi_1(G,\bg) \to \EEx$ using the canonical topological group homomorphism
  $\pi_1(G,\bg) \to \Gal(\bFq/\Fq)$. 
 % 
  The $\Fq$ rational point $g$ also
  determines an equivalence between the category of $\ell$-adic
  representations of $\pi_1(G,\bg)$ and $\ell$-adic local systems on $G$. Let
  $\qcs{L}_0$ be a local system on $G$ in the isomorphism class
  determined by this $\ell$-adic character of $\pi_1(G,\bg)$.
  Then $b_G^*\qcs{E} \iso \gqcs{L}$.

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object of $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function-sheaf dictionary for connected commutative algebraic groups over finite fields, as in \cite{deligne:SGA4.5}*{Sommes trig.} or \cite{laumon:87a}*{1.1.3}, gives us all we need, as we briefly recall. \todo{Fix citelist} %\citelist{\cite{deligne:SGA4.5}*{Sommes trig.},\cite{laumon:87a}*{1.1.3}}
The key observation is that, by \'etale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ ($\Lang$ is proper, so $\Lang_* = \Lang_!$) of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

@@

Although the category of quasicharacter sheaves on $G$ does specialize to that common notion when $G$ is connected and of finite type, the common notion is not appropriate if one wishes to extend the dictionary to smooth commutative group schemes over finite fields. 
In particular, as we have already seen in Section~\ref{sec:etale}, for a given quasicharacter sheaf $\qcs{L} = (\gqcs{L},\mu,\phi)$ there may be a different isomorphism $\mu' : \bm^*\gqcs{L} \to \gqcs{L}\boxtimes\gqcs{L}$ for which $\qcs{L}'=(\gqcs{L},\mu',\phi)$ is also a quasicharacter sheaf on $G$, not isomorphic to $\qcs{L}$ in $\QC(G)$.
Indeed, Proposition~\ref{prop:etale-iso} shows that when $G$ is \'etale, $\Hh^2(\bG,\EEx)^{\Frob{}}$ measures the possibilities for $\mu'$.
In other words, the common perspective on the function-sheaf dictionary leads only, and exactly, to the part of category of quasicharacter sheaves that is {\it completely invisible} to the trace of Frobenius!

We will see in Section~\ref{sec:main} that, for any smooth commutative group scheme $G$, the part of category of quasicharacter sheaves that is invisible to the trace of Frobenius are measured by $\Hh^2(\pi_0(\bG),\EEx)^{\Frob{}}$.
%\end{remark}

@@

\begin{theorem}\label{thm:QCXK}
Let $K$ and $R$ be as above. 
\begin{enumerate}
\item
The trace of Frobenius provides a natural transformation between the additive functors
\[
F_{\QC}: X_K \mapsto \QCiso{X_K}
\qquad\text{and}\qquad
F_{\text{qc}}: X_K \mapsto \Hom_{\text{qc}}(X(K),\EEx)
\]
as functors from $\mathcal{N}_K$ to the category of abelian groups.
\end{enumerate} 
Regarding this natural transformation, for every $X_K \in \mathcal{N}_K$ with N\'eron model $X$:
\begin{enumerate}[resume]
\item there is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{X_K} \to \Hom_\text{qc}(X(K),\EEx) \to 0;
\] 
\item for all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism;
\item for all quasicharacter sheaves $\qcs{F}$ on $\Gr_R(X)$ there is a canonical isomorphism
\[
\Aut(\qcs{F}) \iso \Hom((\pi_0(X)_{\bFq})_{\Weil{}},\EEx).
\]
\end{enumerate}
\end{theorem}

@@

As we explained in the Introduction, our original motivation for this paper was to find a geometrization and categorification of quasicharacters of $p$-adic tori. This is is now provided by specializing Theorem~\ref{thm:QCXK} to the case when $X_K= T\in \mathcal{N}_K$ is an algebraic torus over $K$, in which case we write $\underline{T}$ for the Neron model $X$. Now,
\[
 \pi_0(\underline{T})_{\bFq} = X_*(T)_{\mathcal{I}_K},
\]
where $X_*(T)$ is the cocharacter lattice of $T$ and $\mathcal{I}_K$ is the inertia group for $K$ \cite{bitan:discriminant}*{Eq 3.1}. 
Accordingly, the short exact sequence in Theorem~\ref{thm:QCXK} becomes
\begin{equation}
0 \to \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Frob{}} \to \QCiso{T} \to \Hom_\text{qc}(T(K),\EEx) \to 0.
\end{equation}
Again by Theorem~\ref{thm:QCXK}, the automorphism groups in the $\QC(T)$ are given, for every quasicharacter sheaf $\qcs{F}$ for $T$, by
\begin{equation}
\Aut(\qcs{F}) \iso \Hom(X_*(T)_{\Weil{K}},\EEx) \iso \Hom(X_*(T),\EEx)^{\Weil{K}} = (\check{T}_\ell)^{\Weil{K}}
\end{equation}
where $\Weil{K}$ is the Weil group for $K$ and $\check{T}_\ell$ is the $\ell$-adic dual torus to $T$.

>>>>

  Now let $(\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$
  and let $\rho : \W(G, \bg) \to \EEx$ be the character determined by $(\gqcs{L},\phi)$.
  Composing with the splitting $\Weil{} \to \W(G,\bg)$ yields an $\ell$-adic character
  $\rho_g : \Weil{} \to \EEx$, which is the same as the Trace of Frobenius defined in Section~\ref{ssec:Frob}, for every $\Fq$ rational point $g$ on $G$:
  $
  \rho_g(\Frob{}) =  \trFrob{\qcs{L}}(g).
  $
%
  On the other hand, we have already seen that $\trFrob{\qcs{L}} : G(\Fq) \to \EEx$
  is a group homomorphism. 
  Since $G$ is an algebraic group over $\Fq$, $G(\Fq)$ is finite.
  Therefore $\trFrob{\qcs{L}}(g) = \rho_g(\Frob{})$ is a root of unity
  for every $g\in G(\Fq)$.  Since $\Weil{}$ is generated by
  $\Frob{}$ and $\rho_g : \Weil{} \to \EEx$ is
  a character, it follows that the image of $\rho_g$ is a finite group.
  Thus, $\rho_g$ extends to an $\ell$-adic character of $\Gal(\bFq/\Fq)$,
  which we will also denote $\rho_g$.
%
  We may now lift the $\ell$-adic character $\rho_g : \Gal(\bFq/\Fq) \to \EEx$
  to an $\ell$-adic character $\pi_1(G,\bg) \to \EEx$ using the canonical topological group homomorphism
  $\pi_1(G,\bg) \to \Gal(\bFq/\Fq)$. 
 % 
  The $\Fq$ rational point $g$ also
  determines an equivalence between the category of $\ell$-adic
  representations of $\pi_1(G,\bg)$ and $\ell$-adic local systems on $G$. Let
  $\qcs{E}$ be a local system on $G$ in the isomorphism class
  determined by this $\ell$-adic character of $\pi_1(G,\bg)$.
  Then $b_G^*\qcs{E} \iso \gqcs{L}$.

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object of $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function-sheaf dictionary for connected commutative algebraic groups over finite fields, as in \cite{deligne:SGA4.5}*{Sommes trig.} or \cite{laumon:87a}*{1.1.3}, gives us all we need, as we briefly recall. \todo{Fix citelist} %\citelist{\cite{deligne:SGA4.5}*{Sommes trig.},\cite{laumon:87a}*{1.1.3}}
The key observation is that, by \'etale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

@@

Although the category of quasicharacter sheaves on $G$ does specialize to that common notion when $G$ is connected and of finite type, the common notion is not appropriate if one wishes to extend the dictionary to smooth commutative group schemes over finite fields. 
In particular, as we have already seen in Section~\ref{sec:etale}, for a given quasicharacter sheaf $\qcs{L} = (\gqcs{L},\mu,\phi)$ there may be a different isomorphism $\mu' : \bm^*\gqcs{L} \to \gqcs{L}\boxtimes\gqcs{L}$ for which $\qcs{L}'=(\gqcs{L},\mu',\phi)$ is also a quasicharacter sheaf on $G$, not isomorphic to $\qcs{L}$ in $\QC(G)$.
Indeed, Proposition~\ref{prop:etale-iso} shows that when $G$ is \'etale, $\Hh^2(\bG,\EEx)^{\Weil{}}$ measures the possibilities for $\mu'$.
The common perspective on the function-sheaf dictionary leads only to the part of category of quasicharacter sheaves that is invisible to the trace of Frobenius.

We will see in Section~\ref{sec:main} that, for any smooth commutative group scheme $G$, the part of category of quasicharacter sheaves that is invisible to the trace of Frobenius are measured by $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$.
%\end{remark}

@@

\begin{theorem}\label{thm:QCXK}
Let $K$ and $R$ be as above. 
\begin{enumerate}
\item
The trace of Frobenius provides a natural transformation between the additive functors
\[
F_{\QC}: X_K \mapsto \QCiso{X_K}
\qquad\text{and}\qquad
F_{\text{qc}}: X_K \mapsto \Hom_{\text{qc}}(X(K),\EEx)
\]
as functors from $\mathcal{N}_K$ to the category of abelian groups.
\end{enumerate} 
Regarding this natural transformation, for every $X_K \in \mathcal{N}_K$ with N\'eron model $X$:
\begin{enumerate}[resume]
\item there is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Weil{}} \to \QCiso{X_K} \to \Hom_\text{qc}(X(K),\EEx) \to 0;
\] 
\item for all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism;
\item for all quasicharacter sheaves $\qcs{F}$ on $\Gr_R(X)$ there is a canonical isomorphism
\[
\Aut(\qcs{F}) \iso \Hom((\pi_0(X)_{\bFq})_{\Weil{}},\EEx).
\]
\end{enumerate}
\end{theorem}

@@

As we explained in the Introduction, our original motivation for this paper was to find a geometrization and categorification of quasicharacters of $p$-adic tori. This is is now provided by specializing Theorem~\ref{thm:QCXK} to the case when $X_K= T\in \mathcal{N}_K$ is an algebraic torus over $K$, in which case we write $\underline{T}$ for the Neron model $X$. Now,
\[
 \pi_0(\underline{T})_{\bFq} = X_*(T)_{\mathcal{I}_K},
\]
where $X_*(T)$ is the cocharacter lattice of $T$ and $\mathcal{I}_K$ is the inertia group for $K$ \cite{bitan:discriminant}*{Eq 3.1}. 
Accordingly, the short exact sequence in Theorem~\ref{thm:QCXK} becomes
\begin{equation}
0 \to \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{}} \to \QCiso{T} \to \Hom_\text{qc}(T(K),\EEx) \to 0.
\end{equation}
Again by Theorem~\ref{thm:QCXK}, the automorphism groups in the $\QC(T)$ are given, for every quasicharacter sheaf $\qcs{F}$ for $T$, by
\begin{equation}
\Aut(\qcs{F}) \iso \Hom(X_*(T)_{\Weil{K}},\EEx) \iso \Hom(X_*(T),\EEx)^{\Weil{K}} = (\check{T}_\ell)^{\Weil{K}}
\end{equation}
where $\Weil{K}$ is the Weil group for $K$ and $\check{T}_\ell$ is the $\ell$-adic dual torus to $T$.
